---
Author: Vikas Nelamangala
Date: Saturday, March 20, 2015
Title: Impact of Severe Weather Events on Health and Economy of United States
output: html_document
---

##Synopsis 

In this report we try to answer some basic questions about severe weather events.  Specifically, we try to identify which types of events are the most harmful to population health and the most deleterious to the economy.To answer these questions, we obtained the storm database from the U.S. National Oceanic and Atmospheric Administration's (NOAA).  This database tracks characteristics of major storms and weather events in the United States,including estimates of any fatalities, injuries, and property and crop damage. From this data, we found that tornadoes and heat are the severe weather event types most dangerous to people, whereas flooding, hurricanes, and storm surges are costliest event types affecting the economy.  Interestingly,flooding is one of the top three 'most dangerous' or 'most costly' event.

**About the Data**

The weather events are divided into 13 groups:

-Convection (e.g. tornado, lightning, thunderstorm, hail)
-Flood (e.g. flash flood, river flood)
-Extreme temperatures (e.g. extreme cold, extreme hot)
-Marine (e.g. tsunami, coastal storm, rip current, high waves, high seas)
-Winter (e.g. avalanche, snow, blizzard, icy roads, freeze)
-Tropical Cyclones (e.g. tropical storm, hurricane)
-High Wind (e.g. winds, microburst)
-Fire
-Rain
-Drought/Dust (e.g. drought, dust storm, dust)
-Landslide
-Fog
-Others

## Data Processing

```{r importdata, echo=TRUE}
#Setting WD
setwd("~/Desktop/Coursera/ReproResearch/my/RepData_PeerAssessment2")

#Unzip and read .csv file into the variable data
unzip <- bzfile("repdata-data-StormData.csv.bz2", "r")
data <- read.csv(unzip, stringsAsFactors = FALSE)
close(unzip)
```

**Select useful data** 

Subsetting data into variables that are needed and adding a new variable.

```{r datasubset,echo=TRUE,cache=TRUE}
x <- which(colnames(data) %in% c("BGN_DATE", "PROPDMG", "CROPDMG", "EVTYPE", 
    "INJURIES", "FATALITIES"))
data <- data[, x]
head(data)

#Formatting date and time
data$YEAR <- as.integer(format(as.Date(data$BGN_DATE, "%m/%d/%Y 0:00:00"), "%Y"))
head(data)

#To uppercase
data$EVTYPE <- toupper(data$EVTYPE)
head(data)

# creates new variable
data$ECONOMICDMG <- data$PROPDMG + data$CROPDMG
head(data)

# Select only positive value data
data <- subset(data, data$FATALITIES > 0 | data$ECONOMICDMG > 0 | data$INJURIES > 
    0)
head(data)
```

Data aggregation

```{r aggregate,echo=TRUE,cache=TRUE}
library(plyr)

# data aggregated by YEAR & EVTYPE.
#ddply -> For each subset of a data frame, apply function then combine results into a data frame.

eventYear <- ddply(data[, -1], .(YEAR, EVTYPE),
                   .fun = function(x) {
                         return(
                           c(sum(x$FATALITIES), sum(x$ECONOMICDMG), sum(x$INJURIES))
                              )
                                      }
                   )
names(eventYear) <- c("YEAR", "EVTYPE", "FATALITIES", "ECONOMICDMG", "INJURIES")
head(eventYear)
```

**Grouping the events**
We grouped the events by its related categories

```{r group,echo=TRUE,cache=TRUE}
#Function that calculates the events by categories (13 categories described in the synopsis)

#grepl -> search for matches to argument pattern within each element of a character vector

eventCategory <- function(x) {
    ev <- x$EVTYPE[1]
    if (grepl("LIG(H|N)T(N|)ING|TORNADO|T(H|)U(N|)(DER|ER|DEER|DERE)(STORM|STROM|TORM)|TSTM|HAIL", 
        ev)) {
        category <- "Convection"
    } else if (grepl("WINT(ER|RY)|ICE|AVALANC(H|)E|SNOW|BLIZZARD|FREEZ|ICY|FROST", 
        ev)) {
        category <- "Winter"
    } else if (grepl("COLD|HEAT|HOT|TEMPERATURE|COOL|WARM", ev)) {
        category <- "Extreme Temp"
    } else if (grepl("FLOOD| FLD$", ev)) {
        category <- "Flood"
    } else if (grepl("COASTAL|TSUNAMI|RIP CURRENT|MARINE|WATERSPOUT|SURF|SLEET|SEAS|(HIGH|RISING|HEAVY) (WAVES|SWELLS|WATER)", 
        ev)) {
        category <- "Marine"
    } else if (grepl("TROPICAL|HURRICANE|STORM SURGE|TYPHOON", ev)) {
        category <- "Tropical Cyclones"
    } else if (grepl("WIND|MICROBURST", ev)) {
        category <- "High Wind"
    } else if (grepl("FIRE", ev)) {
        category <- "Fire"
    } else if (grepl("RAIN|PRECIP", ev)) {
        category <- "Rain"
    } else if (grepl("DROUGHT|DUST", ev)) {
        category <- "Drought/Dust"
    } else if (grepl("LANDSLIDE|MUD.*SLIDE", ev)) {
        category <- "Landslide"
    } else if (grepl("FOG|VOG", ev)) {
        category <- "Fog"
    } else {
        category <- "Others"
    }

    x$EVGROUP <- rep(category, dim(x)[1])
    return(x)
}
eventYear <- ddply(eventYear, .(EVTYPE), .fun = eventCategory)
head(eventYear)

#We organize the data to show FATALITIES, ECONOMICDMG and INJURIES
#by YEAR and EVGROUP

groupYear <- ddply(eventYear, .(YEAR, EVGROUP), .fun = function(x) {
    return(c(sum(x$FATALITIES), sum(x$ECONOMICDMG), sum(x$INJURIES)))
})

names(groupYear) <- c("YEAR", "EVGROUP", "FATALITIES", "ECONOMICDMG", "INJURIES")
head(groupYear)

# calculate average annual damage by group
eventFirstYear <- ddply(groupYear, .(EVGROUP), .fun = function(x) {
    return(c(min(x$YEAR)))
})
names(eventFirstYear) <- c("Weather.Event", "First.Year")
head(eventFirstYear)
```

As we can notice analysing the variable eventFirstYear, the weather event "Convection" has its occurency starting at the 50's but the others events starts at 1993. In this section we subset the groupYear to analysis all the events starting from 1993

```{r data1993,echo=TRUE,cache=TRUE}

## start data analysis at 1993
groupYear <- subset(groupYear, YEAR >= 1993)

# calculate average annual damage by group
byGroup <- ddply(groupYear, .(EVGROUP), .fun = function(x) {
    return(c(mean(x$FATALITIES), mean(x$ECONOMICDMG), mean(x$INJURIES)))
})
names(byGroup) <- c("EVGROUP", "AVG.FATALITIES", "AVG.ECONOMICDMG", "AVG.INJURIES")
head(byGroup)

```

## Results

**Results section 1 - Health Harmful Events**

This histograms Show fatalities and injuries for weather events.

```{r results1,echo=TRUE, cache=TRUE}
# Graph libraries
library(ggplot2)
library(scales)

# average annual populational damage by group of event
byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.FATALITIES))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_histogram(aes(weight = AVG.FATALITIES, fill = AVG.FATALITIES), binwidth = 5, 
    color = "black") + ggtitle("Average Fatalities") + ylab("# Fatalities") + 
    xlab("Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))

# average annual populational damage by group of event
byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.INJURIES))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_histogram(aes(weight = AVG.INJURIES, fill = AVG.INJURIES), binwidth = 1, 
    color = "black") + ggtitle("Average Injuries") + ylab("# Injuries") + xlab("Weather Event") + 
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


**Results section 2 - Economic Harm**
  
Histogram of weather event harm to the economy.

```{r results2,echo=TRUE, cache=TRUE}
# average annual economical damage by group of event
byGroup$EVGROUP <- with(byGroup, reorder(EVGROUP, -AVG.ECONOMICDMG))
g <- ggplot(byGroup, aes(x = EVGROUP))
g + geom_histogram(aes(weight = AVG.ECONOMICDMG, fill = AVG.ECONOMICDMG), binwidth = 1, 
    color = "black") + ggtitle("Average Economic Damage") + ylab("Economic damage") + 
    xlab("Weather Event") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Conclusion 
From this plot, we can see that the most impactful event types are either very dangerous to the population or very costly, but generally not at the same time. For example, heat injures and kills many people but is not particularly costly. Similarly, storm surge is costly but is relatively harmless to people.

According to the analysis, we can notice in the results that the most harmful events for population are "Extreme temperatures" and "Convection" when we look at "Average Fatalities". When we talk about "Average Injuries", we have the same events, but in a different order - "Convection" and "Extreme Temperatures". Now, when we look at economic damage,the extremely harmful events for economy are "Convection" and "Flood". 

Tornadoes are the most dangerous to people but storm surges, hurricanes, and floods are most costly.  Also, flooding is the most costly, but heat and tornadoes injure or kill more people.